---
title: "Hulle_Madeline_Individual_Assignment_05"
author: "Madeline Hulle"
format: pdf
date: "2026-05-01"
execute: 
  message: false
  warning: false
---

# Set Up

## Packages
```{r}
#| label: reading-packages

# general use
library("tidyverse")
library("janitor")

# file organization
library("here")

# data visualization
library("wesanderson")

```

## Data
```{r}
#| label: reading-data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

#survey data (reading different sheets in as seperate data frames)

aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")

# site information

sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quaity

water_quality <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "Water Quality",
   # values in df that are actually indicating missing values, making it seem as a chr when it should be num
   # different ways missing values show up in data set
   na = c("over", "", "n/a", "N/A", "173+")
)

```

# Class Code

# 2. Cleaning

## `sites` object

```{r}
#| label: cleaning-sites

# creating new object from sites
ncos_sites <- sites |> 
  # filtering to only include NCOS quarterly sites
  filter(sector == "NCOS" & 
           sampling_frequency == "Quarterly")

```

## clean `taxon_list`


```{r}
#| label: taxon-list-cleaning

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converted taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))


```


## clean `water_quality`

```{r}
#| label: cleaning-water-quality

# creatign new object from water_quality
water_quality_clean <- water_quality |> 
  # cleaning column name
  clean_names() |> 
  # a filtering join, only keep the rows in wq clean in which the site matches up with the site in ncos clean
  semi_join(ncos_sites, by = "site") |> 
  # creating new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # grouping by date/site
  group_by(date_site) |> 
  # calculating the medians
  summarize(med_pH = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungrouping df
  ungroup() |> 
  # filtering out outliers
  filter(date_site != "2022-08-12_NVBR")



```

## clean `aq_ins`

```{r}
#| label: cleaning_aquatic_insects

# create a new object
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join only include quarterly ncos surveys
  semi_join(ncos_sites, by = "site") |> 
  # renaming date_on_vial column to date
  rename(date = date_on_vial) |> 
  # selecting columns of interest
  # keeping information about each survey and the 9 most abundant taxa
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filtered to only include filteres beaker samples (better sampling technique)
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # converting to long format, making df longer 
  pivot_longer(
    # name the columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for old column names
    # column names are all in new column called taxon
    names_to = "taxon",
    # new column for the values in each of the old columns
    # counts now under a column called abundance
    values_to = "abundance"
  ) |> 
  # group by date site and taxon
  group_by(date, site, taxon) |> 
  # calculated average count per liter (total abundance/7.7)
  summarise(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungrouping data frame
  ungroup() |> 
  # new column to join data frames, didnt remove them 
  unite("date_site", date, site, remove = FALSE) |> 
  # joined with water quality clean
  left_join(water_quality_clean, by = "date_site") |> 
  # joined with taxanomic information
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # fill in full site names 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "East Channel",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # set factor levels for full site name by median salinity
  mutate(site_full = fct_reorder(site_full,
                                  med_sal,
                                  .fun = "median",
                                  .na_rm = TRUE),
         # reversing the order 
         site_full = fct_rev(site_full))






