---
title: "Hulle_Madeline_Individual_Assignment_05"
author: "Madeline Hulle"
format: pdf
date: "2026-05-09"
execute: 
  message: false
  warning: false
---

# Set Up

## Packages
```{r}
#| label: reading-packages

# general use and cleaning

library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)

# coloring
library(wesanderson)


```

## Data
```{r}
#| label: reading-data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

#survey data (reading different sheets in as seperate data frames)

aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")

# site information

sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quaity

water_quality <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "Water Quality",
   # values in df that are actually indicating missing values, 
   # making it seem as a chr when it should be num
   # different ways missing values show up in data set
   na = c("over", "", "n/a", "N/A", "173+")
)

```

# Class Code

## 2. Cleaning

### `sites` object

```{r}
#| label: cleaning-sites

# creating new object from sites
ncos_sites <- sites |> 
  # filtering to only include NCOS quarterly sites
  filter(sector == "NCOS" & 
           sampling_frequency == "Quarterly")

```

### clean `taxon_list`


```{r}
#| label: taxon-list-cleaning

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converted taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))


```


### clean `water_quality`

```{r}
#| label: cleaning-water-quality

# creatign new object from water_quality
water_quality_clean <- water_quality |> 
  # cleaning column name
  clean_names() |> 
  # a filtering join, only keep the rows in wq clean in which the 
  # site matches up with the site in ncos clean
  semi_join(ncos_sites, by = "site") |> 
  # creating new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # grouping by date/site
  group_by(date_site) |> 
  # calculating the medians
  summarize(med_pH = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungrouping df
  ungroup() |> 
  # filtering out outliers
  filter(date_site != "2022-08-12_NVBR")



```

### clean `aq_ins`

```{r}
#| label: cleaning_aquatic_insects

# create a new object
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join only include quarterly ncos surveys
  semi_join(ncos_sites, by = "site") |> 
  # renaming date_on_vial column to date
  rename(date = date_on_vial) |> 
  # selecting columns of interest
  # keeping information about each survey and the 9 most abundant taxa
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filtered to only include filteres beaker samples 
  # (better sampling technique)
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # converting to long format, making df longer 
  pivot_longer(
    # name the columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for old column names
    # column names are all in new column called taxon
    names_to = "taxon",
    # new column for the values in each of the old columns
    # counts now under a column called abundance
    values_to = "abundance"
  ) |> 
  # group by date site and taxon
  group_by(date, site, taxon) |> 
  # calculated average count per liter (total abundance/7.7)
  summarise(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungrouping data frame
  ungroup() |> 
  # new column to join data frames, didnt remove them 
  unite("date_site", date, site, remove = FALSE) |> 
  # joined with water quality clean
  left_join(water_quality_clean, by = "date_site") |> 
  # joined with taxanomic information
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # fill in full site names 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "East Channel",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # set factor levels for full site name by median salinity
  mutate(site_full = fct_reorder(site_full,
                                  med_sal,
                                  .fun = "median",
                                  .na_rm = TRUE),
         # reversing the order 
         site_full = fct_rev(site_full))

``` 

## Visualizing

### Salinity
```{r}
#| label: salinity-plot
#| fig-height: 6
#| fig-width: 8

# creating new objcect to store plot 
salinity_plot <- ggplot(data = aq_ins_clean,
                        mapping = aes(x = site_full,
                                      y = med_sal)) +
  # points representing median salinity at each survey                        
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # points representing mean salinity
  stat_summary(geom = "point",
               fun = median,
               color = "red",
               size = 3) +
  # adding line breaks to x axis labels (wrapping)
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # relabeling axes and adding a title
  labs(x = "Site",
       y = "Median Salinity (ppt)",
       title = "Site Salinity") +
  # adding themeing
  theme_bw()

```

### Copepod Abundance

#### Cleaning
```{r}
#| label: copepod-cleaning

copepods <- aq_ins_clean |> 
  filter(taxon == "copepod") |> 
  # adding 1 bc some observations are 0
  # helps overcome issues with large differences in order of magnitude
  mutate(log_ave_lit = log(ave_lit + 1))




```

#### Copepod Viz
```{r}
#| label: copepod-plot
#| fig-height: 6
#| fig-width: 8

# creating object to store plot
copepod_plot <- ggplot(data = copepods,
                       mapping = aes(x = site_full,
                                     # using log transformed count per liter
                                     y = log_ave_lit)) +
  # jitter to represent each sample
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # points to represent median copepod abundance
  stat_summary(geom = "point",
               fun = median,
               color = "blue",
               size = 3) +
  # wrappign x-axis labels
  scale_x_discrete(labels = label_wrap(width = 10)) +
  # relabeling axes 
  labs(x = "Site",
       y = "Log +1 transformed average abundance/L ",
       title = "Copepod Abundance") +
  # using different theme
  theme_bw()

```

# Problems

## 1. Visualizing differences across sites in major taxon abundance

### a. Plan your figure

- x-axis: `site_full`
- y-axis: `log_ave_lit` (updated w/ `log_ave_lit` after viz)
- geometry to represent average abundance/L: geom_jitter()
- geometry to represent medians: stat_summary(geom = "point", fun = median)

### b. Wrangle your data
```{r}
#| label: ostracod-wrangling

# creating a new object `ostracods` from `aq_ins_clean`
ostracods <- aq_ins_clean |> 
  # filtering to only include ostracods
  filter(taxon == "ostracod") |> 
  # log transforming `ave_lit` to overcome large difference in
  # order of magnitude between observations
  mutate(log_ave_lit = log(ave_lit + 1))

# taking a sample of 10 observations
slice_sample(ostracods, n = 10)

```

### c. Code Your Figure
```{r}
#| label: ostracod-viz
#| fig-height: 6
#| fig-width: 8

ostracod_plot <- ggplot(data = ostracods,
                        mapping = aes(x = site_full,
                                      y = log_ave_lit)) +
  geom_jitter(height = 0,
              width = 0.2,
              shape = 21) +
  stat_summary(geom = "point",
               fun = median,
               color = "hotpink",
               size = 3) +
  labs(x = "Site",
       y = "Log +1 Transformed Average Abundance/Liter",
       title = "Ostracod Abundance") +
  theme_bw() +
  scale_x_discrete(labels = label_wrap(width = 10))

ostracod_plot

```


### d. Create a Multi Panel Figure
```{r}
#| label: multi-panel-viz
#| fig-height: 8
#| fig-width: 10


salinity_plot / (copepod_plot | ostracod_plot)

```

### e. Interpret Your Figure

- Between sites there does seem to be a difference between the median log + 1 average abundance/L of Copepods (blue dots) and that of Ostracods (pink dots). For sites North of Venoco Bridge and East Bridge, the median log + 1 average abundance/L of Ostracods is zero while for Copepods, this value is greater than zero, with East Channel being the site with the greatest median log + 1 average abundance/L of all of the sites. For the Main Channel site, Ostracod and Copepod median log + 1 average abundance/L are quite similar, both slightly above 0. The site South of Phelps Bridge also sees similar levels of median log + 1 average abundance/L across both groups, with this being the highest median log + 1 average abundance/L seen for the Ostracods. At North of Phelps Bridge we see low median log + 1 average abundance/L for both groups, with Ostracod having a slightly lower median log + 1 average abundance/L than Copepods. Finally, Phelps Road and Devereux Creek see similar relative median log + 1 average abundance/L values across the two groups, but it is higher for Copepods at both sites. Overall, there seems to be, on average, higher Copepod abundance across these 7 sites than Ostracod abundance as more of the blue (median log + 1 average abundance/L for Copepods) points are higher up on the y-axis than the pink ones (median log + 1 average abundance/L for Ostracods).

- Comparing the Site Salinity plot to the Ostracod Abundance plot, there seems to be a possible relationship between the two. For site salinity, the mean salinity level is highest for North of Venoco Bridge and steadily decreases through the intermediate sites until the lowest median salinity is recorded at Devereux Creek. In the Ostracod Abundance plot, we see an almost inverse pattern where North of Venoco Bridge and East Channel see the lower median log + 1 average abundance/L levels and this steadily increases (with the exception of the South of Phelps Bridge observation) along the intermediate sites and reaching a peak at Devereux creek. More data would need to be collected and more analysis be done to determien if this is a true relationship and if the South of Phelps Bridge observation is a fluke) or if these inverse shapes are a coincidence in the data. From my visualization now however, it seems possible that higher salinity corresponds with a lower abundance of Ostracods and a lower salinity corresponds to a higher abundance of Ostracods. The optimal environmental conditions of Ostracods would also be helpful information to determine the validity of this relationship.

## 2.  Visualizing total abundance as a function of salinity

### a. Plan Your Figure

- x-axis: `med_sal`
- y-axis: `log_ave_lit`
- color: taxon
- geometry: geom_jitter()
- facets: facet_wrap(~taxon, scales = "free")

### b. Wrangle Your Data
```{r}
#| label: abundance-salinity-wrangle

# creating a new object `log_aq_ins_clean` from `aq_ins_clean`
log_aq_ins_clean <- aq_ins_clean |> 
  # log transforming the `ave_lit` column
  mutate(log_ave_lit = log(ave_lit + 1)) |> 
  # turnign each taxon into title case
  mutate(taxon = to_any_case(taxon,
                             case = "title")) |> 
  # reordering the levels of taxon by the max abundance
  mutate(taxon  = fct_rev(fct_reorder(taxon,
                              ave_lit,
                              .fun = max,
                              na.rm = TRUE)))

# taking a sample of 10 observations
slice_sample(ostracods, n = 10)

```

### c. Code Your Figure
```{r}
#| label: abundance-salinity-viz
#| fig-height: 6
#| fig-width: 8

# creating a new object to store plot from the data `log_aq_ins_clean
log_abundance_salinity <- ggplot(data = log_aq_ins_clean,
                                 # x-axis: median salinity
                                 # y-axis: log transformed average abundance/L
                                 mapping = aes(x = med_sal,
                                               y = log_ave_lit,
                                               # coloring by taxon
                                               color = taxon)) +
  # points representing median salinity 
  geom_jitter(height = 0) +
  # plot for each taxon with own y-axis 
  facet_wrap(~ taxon,
             scales = "free") +
  # adding in custom coloring
  scale_color_manual(values = wes_palette("GrandBudapest2",
                                          n = 9, type = "continuous")) +
  # custom theme
  theme_bw() +
  # taking out legend, redundant
  theme(legend.position = "none") +
  # labeling x and y axes 
  labs(x = "Median Salinity (ppt)",
       y = "Log + 1 Transformed Average Abundance/Liter",
       # adding a title
       title = "Abundance as a Function of Salinity")

# displaying the plot
log_abundance_salinity

```

### d. Interpret Your Figure

Ostracod, Copepod, and Cladocera show the highest log-transformed abundances overall (y-axes reaching 6–7), and all three display a clear negative relationship with salinity as observations cluster heavily at low salinity (0–10 ppt) with higher abundance values, a pattern that decreases with increased salinity values. Hemiptera Corixidae Boatman follows a similar pattern of low-salinity concentration but at much lower maximum abundances (~4). The annelid taxa (Polychaete and Oligochaete) and the dipteran taxa (Chironomid and Ceratopogonidae) are present across the full salinity range (0–70 ppt), suggesting stronger salinity tolerance, though their abundances remain consistently low regardless of salinity. Nematode stands out as the taxon with the lowest maximum abundance of all nine (~0.25), with observations scattered across salinity levels and no discernible trend. Overall, most taxa appear to prefer low-salinity conditions, but the top three taxa, Ostracod, Copepod, Cladocera, show this preference most dramatically in both abundance magnitude and distributional pattern.


