---
title: "In-Class Template"
author: "Madeline Hulle"
format: pdf
---

# 1. Set Up

## Packages
```{r}
#| label: packages

# general use and cleaning

library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualize mising data
library(naniar)

# arrange plots
library(patchwork)


```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

#survey data (reading different sheets in as seperate data frames)

aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")

# site information

sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quaity

water_quality <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "Water Quality",
   # values in df that are actually indicating missing values, making it seem as a chr when it should be num
   # different ways missing values show up in data set
   na = c("over", "", "n/a", "N/A", "173+")
)



```


# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` that only includes NCOS sites

```{r}
#| label: cleaning-sites

# creating new oject from sites
ncos_sites <- sites |> 
  # filtering to only include NCOS quarterly sites
  filter(sector == "NCOS" & 
           sampling_frequency == "Quarterly")


```

## clean `taxn_list`

Goal: convert chr strings in the existing `taxon_list` df into snakecase

```{r}
#| label: taxon-list-cleaning

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # coverted taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))




```


## clean `water_quality`

Goal: to create a df with median pH, median DO (mg/L), median salinity (ppt) for each data and site

- create new object from `water_quality`
- clean column names
- only include surveys from quarterly NCOS sites
- create a new column called `date_site` to join with later data frames
- group_by date_site
- calculate median pH, salinity, and DO
-ungroup data frame
-filter out any outliers



```{r}
#| label: cleaning-water-quality

# creatign new object from water_quality
water_quality_clean <- water_quality |> 
  # cleaning column name
  clean_names() |> 
  # a filtering join, only keep the rows in wq clean in which the site matches up with the site in ncos clean
  semi_join(ncos_sites, by = "site") |> 
  # creating new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # grouping by date/site
  group_by(date_site) |> 
  # calculating the medians
  summarize(med_pH = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungrouping df
  ungroup() |> 
  # filtering out outliers
  filter(date_site != "2022-08-12_NVBR")



```

## clean `aq_ins`

Goal: create a long format data frame (currently in wide format, like lots of other biological data sets) in which each row contains the counted indoviduals in each taxon during each survey. The data frame should include taxanomic information for each organism and the water quality information for each survey.

*new df should have more rows*

*functions in R expect long format*

- create a new object from `aq_ins`
- clean column names
- filter tp only include quarterly NCOS sites
- select columns
- filter to only include "filtered beaker" sample
- convert df to long format
- group by ste date and taxon
- calculate avg abundance per liter (sum counts divide dby 7.5 liters)
- create a new column called `date_site`to join with water quality
- join data frames
- join with taxon lists

```{r}
#| label: cleaning_aquatic_insects


aq_ins_clean <- aq_ins |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  rename(date = date_on_vial) |> 
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # converting to long format
  pivot_longer(
    # name the columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for old column names
    # column names are all in new column called taxon
    names_to = "taxon",
    # new column for the values in each of the old columns
    # counts now under a column called abundance
    values_to = "abundance"
  ) |> 
  group_by(date, site, taxon) |> 
  summarise(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  ungroup() |> 
  unite("date_site", date, site, remove = FALSE) |> 
  left_join(water_quality_clean, by = "date_site") |> 
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) 
  
  
  
  
  # wide to long, making missing values explicit







```












