---
title: "In-Class Template"
author: "Madeline Hulle"
format: pdf
execute: 
  message: false
  warning: false
---

# 1. Set Up

## Packages
```{r}
#| label: packages

# general use and cleaning

library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)


```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

#survey data (reading different sheets in as seperate data frames)
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")

# site information
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
   here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
   sheet = "Water Quality",
   # values in df that are actually indicating missing values, making it seem as a chr when it should be num
   # different ways missing values show up in data set
   na = c("over", "", "n/a", "N/A", "173+")
)



```

# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` that only includes NCOS sites

```{r}
#| label: cleaning-sites

# creating new object from sites
ncos_sites <- sites |> 
  # filtering to only include NCOS quarterly sites
  filter(sector == "NCOS" & 
           sampling_frequency == "Quarterly")

```

## clean `taxn_list`

Goal: convert chr strings in the existing `taxon_list` df into snakecase

```{r}
#| label: taxon-list-cleaning

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converted taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))


```


## clean `water_quality`

Goal: to create a df with median pH, median DO (mg/L), median salinity (ppt) for each data and site

- create new object from `water_quality`
- clean column names
- only include surveys from quarterly NCOS sites
- create a new column called `date_site` to join with later data frames
- group_by date_site
- calculate median pH, salinity, and DO
-ungroup data frame
-filter out any outliers

```{r}
#| label: cleaning-water-quality

# creatign new object from water_quality
water_quality_clean <- water_quality |> 
  # cleaning column name
  clean_names() |> 
  # a filtering join, only keep the rows in wq clean in which the site matches up with the site in ncos clean
  semi_join(ncos_sites, by = "site") |> 
  # creating new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # grouping by date/site
  group_by(date_site) |> 
  # calculating the medians
  summarize(med_pH = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungrouping df
  ungroup() |> 
  # filtering out outliers
  filter(date_site != "2022-08-12_NVBR")



```

## clean `aq_ins`

Goal: create a long format data frame (currently in wide format, like lots of other biological data sets) in which each row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information for each organism and the water quality information for each survey.

*new df should have more rows*

*functions in R expect long format*

- create a new object from `aq_ins`
- clean column names
- filter tp only include quarterly NCOS sites
- select columns
- filter to only include "filtered beaker" sample
- convert df to long format
- group by ste date and taxon
- calculate avg abundance per liter (sum counts divide dby 7.5 liters)
- create a new column called `date_site`to join with water quality
- join data frames
- join with taxon lists

```{r}
#| label: cleaning_aquatic_insects

# create a new object
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join only include quarterly ncos surveys
  semi_join(ncos_sites, by = "site") |> 
  # renaming date_on_vial column to date
  rename(date = date_on_vial) |> 
  # selecting columns of interest
  # keeping information about each survey and the 9 most abundant taxa
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filtered to only include filteres beaker samples (better sampling technique)
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # converting to long format, making df longer 
  pivot_longer(
    # name the columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for old column names
    # column names are all in new column called taxon
    names_to = "taxon",
    # new column for the values in each of the old columns
    # counts now under a column called abundance
    values_to = "abundance"
  ) |> 
  # group by date site and taxon
  group_by(date, site, taxon) |> 
  # calculated average count per liter (total abundance/7.7)
  summarise(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungrouping data frame
  ungroup() |> 
  # new column to join data frames, didnt remove them 
  unite("date_site", date, site, remove = FALSE) |> 
  # joined with water quality clean
  left_join(water_quality_clean, by = "date_site") |> 
  # joined with taxanomic information
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # fill in full site names 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "East Channel",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # set factor levels for full site name by median salinity
  mutate(site_full = fct_reorder(site_full,
                                  med_sal,
                                  .fun = "median",
                                  .na_rm = TRUE),
         # reversing the order 
         site_full = fct_rev(site_full))
  
  
  
  
  # wide to long, making missing values explicit


```

# 3. Vusualizing

## Visualize Missing Values
```{r}
#| label: missing-viz

gg_miss_var(aq_ins_clean)


```


Missing values for salinity, pH, and DO might influence later figures

## Visulaize Salinity

- x: axis: site
- y:axis: mean salinity
- geometry tp show salinity durin survey: geom_jitter()
-geometry t show median salinity: stat_summary


```{r}
#| label: salinity-plot
#| fig-height: 6
#| fig-width: 8

# creating new objcect to store plot 
salinity_plot <- ggplot(data = aq_ins_clean,
                        mapping = aes(x = site_full,
                                      y = med_sal)) +
  # points representing median salinity at each survey                        
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # points representing mean salinity
  stat_summary(geom = "point",
               fun = median,
               color = "red",
               size = 4) +
  # adding line breaks to x axis labels (wrapping)
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # relabeling axes and adding a title
  labs(x = "Site",
       y = "Median Salinity (ppt)",
       title = "Site Salinity") +
  # adding themeing
  theme_bw()

# visualizing the object
salinity_plot

```

Sites seem to differ in salinity. North of Venoco bridge, East Channel, and Main Channel tend to be hyper-saline while North of Phelps Bridge, Phelps Road, and Deveroux Creek is the least saline. South of Phelps bridge tends to be more freash water but it higher in salinity than the other, lower salinity sites


## Visualize differences in copepod abundance

- x-axis: site_full
- y-axis: average_abundance_l
- geometry to represent observations: geom_jitter
- geometry to represent medians: stat_summary

### Initial Cleaning

```{r}
#| label: copepod-cleaning

copepods <- aq_ins_clean |> 
  filter(taxon == "copepod") |> 
  # adding 1 bc some observations are 0
  # helps overcome issues with large differences in order of magnitude
  mutate(log_ave_lit = log(ave_lit + 1))




```

### Visualizing
```{r}
#| label: copepod-plot
#| fig-height: 6
#| fig-width: 8

# creating object to store plot
copepod_plot <- ggplot(data = copepods,
                       mapping = aes(x = site_full,
                                     # using log transformed count per liter
                                     y = log_ave_lit)) +
  # jitter to represent each sample
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # points to represent median copepod abundance
  stat_summary(geom = "point",
               fun = median,
               color = "blue",
               size = 2) +
  # wrappign x-axis labels
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # relabeling axes 
  labs(x = "Site",
       y = "log +1 transformed average abundance/L ",
       title = "Copepod Abundance") +
  # using different theme
  theme_bw()

# displaying the object
copepod_plot


```

Copepod abundance differs across sites. Main Channel, North of Phelps tend to be low N of Venoco bridge east channel,...

## Putting Salinity and Copepod abundance together

```{r}
#| label: multi-panel
#| fig-height: 6
#| fig-width: 8

salinity_plot / copepod_plot


```







